---
title: "Assignment 2: Physical Properties of Lakes"
author: "Andrew Friedman-Herring"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Water Data Analytics on the physical properties of lakes. 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After completing your assignment, fill out the assignment completion survey in Sakai. 

Having trouble? See the assignment's answer key if you need a hint. Please try to complete the assignment without the key as much as possible - this is where the learning happens!

Target due date: 2022-01-25

## Setup

-1. Verify your working directory is set to the R project file, 
-2. Load the tidyverse, lubridate, and rLakeAnalyzer packages   
-3. Import the NTL-LTER physical lake dataset and set the date column to the date format.
4. Using the `mutate` function, add a column called Month. Remove temperature NAs.
-5. Set your ggplot theme (can be theme_classic or something else)

```{r}
getwd()
library(tidyverse)
library(lubridate)
library(rLakeAnalyzer)
theme_set(theme_classic())
NTL.data <- read.csv("../Data/Raw/NTL-LTER_Lake_ChemistryPhysics_Raw.csv")
NTL.data$sampledate<- as.Date(NTL.data$sampledate, format = "%m/%d/%y") #changes to date object

#this chunk drops the NAs and adds a month column
NTL.data.clean <- 
  NTL.data %>%
  drop_na(temperature_C)%>%
  mutate(month = month(sampledate))

```

## Creating and analyzing lake temperature profiles

5. For the year 1993, plot temperature and dissolved oxygen profiles for all six lakes in the dataset (as two separate ggplots). Use the `facet_wrap` function to plot each lake as a separate panel in the plot. Plot day of year as your color aesthetic and use a reverse y scale to represent depth. 

What seasonal trends do you observe, and do these manifest differently in each lake?

> Answer: As the day number increases, the temperature of the lake increases, with the top of the lake heating
faster, and the bottom of the lakes increasing slower. Paul lake has a higher temperature in deeper sections of the lake compared to East Long Lake or Tuesday Lake. 
In terms of Dissolved oxygen, it seems that the concentration decreases throughout the summer, and is at the
highest concentration early in the spring. Some lakes, like Paul, Peter, and West Long Lake show large bumps in dissolved oxygen at around 4 meters, whereas Tuesday Lake and East Long Lake do not show the same bump. This is likely due to differences in primary productivity and light penetration differences. 

```{r}
NTL.data.clean.1993 <- NTL.data.clean%>%
  filter(year4 == 1993)

ggplot(NTL.data.clean.1993, aes(x = temperature_C, y = depth, color = daynum))+
  scale_y_reverse()+
  geom_point(alpha = 0.5)+
  scale_color_viridis_c()+
  facet_wrap(~lakename)+
  labs(x = "Temperature *C", y = "Depth", color = "DOY")
  
  

ggplot(NTL.data.clean.1993, aes(x = dissolvedOxygen, y = depth, color = daynum))+
  scale_y_reverse()+
  geom_point(alpha = 0.5)+
  scale_color_viridis_c()+
  facet_wrap(~lakename)+
  labs(x = "Dissolved Oxygen", y = "Depth", color = "DOY" )
  
  
```

6. Create a new dataset that calculates thermocline depths for all lakes on all dates (hint: you will need group by lake, year, month, DOY, and sample date).

```{r}
NTL.data.clean.thermocline <- NTL.data.clean%>%
  group_by(lakename, year4, month, daynum, sampledate)%>%
  summarise(thermocline = thermo.depth(wtr = temperature_C, depths = depth, seasonal = FALSE))%>%
  filter(thermocline>1)
```


7. Plot thermocline depth by day of year for your newly made dataset. Color each point by lake name, make the points 50% transparent, and choose a color palette other than the ggplot default. 

```{r}
ggplot(NTL.data.clean.thermocline, aes(x = daynum, y = thermocline, color = lakename))+
  geom_point(alpha = 0.5)+
  scale_color_viridis_d()+
  labs(x = "DOY", y = "Thermocline Depth", color = "Lake Name")

```

8. Create a boxplot of thermocline depth distributions split up by lake name on the x axis and by month as the fill color (hint: you will need to set Month as a factor). Choose a color palette other than the ggplot default, relabel axes and legend, and place the legend on the top of the graph.

```{r}
NTL.data.clean.thermocline$month<- as.factor(NTL.data.clean.thermocline$month)

ggplot(NTL.data.clean.thermocline, aes(lakename, thermocline, fill = month))+
  geom_boxplot()+
  scale_y_reverse()+
  scale_fill_viridis_d()+
  theme(legend.position = "top")+
  labs(x = "Lake", y = "Thermocline Depth", color = "Month")


```

Do all the lakes have a similar seasonal progression of thermocline deepening? Which lakes have the deepest thermoclines, and how does this relate to their maximum depth? 

> Answer: Yes, all lakes show at least some level of thermocline deepening throughout the year. 
The magnitude of the depening is highly variable, with Hummingbird Lake having a very 
shallow thermoclline with not much variation, whereas Crampton, Peter and Paul Lakes 
all have very deep thermoclines, with a dramatic progression through the seasons. 
In general, the lakes that were the deepest showed a larger change in thermocline depth, in 
addition to deeper thermocline values in general.


